### 2023-04-25 Manuscript/Draft1.Rmd
a few manuscript updates

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output. While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used. As a result, although $132/160=82.5\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.

@@

@@

Of the scenarios without covariates, $10/16\approx63$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates). Adding covariate information resulted in $14/16\approx88$\% of VAST estimates having improved error compared to the stratified mean. The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario. Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.

Generally help, but dependent on appropriate covariate response. Yellowtail cases always had appropriate response and thus adding covariates always improved estimates. VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed. (show plot comparing good vs bad response)

@@

**For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research**

@@

<div align="center">Table XX. Parameters used in all VAST models.</div>

@@

tabl <- "
|	Parameter	|	Description | Settings A | Settings B
|:--	|:------- |:---- |:----
| *ObsModel* | Link function and assumed distribution | c(10,2) | c(10,2)
|FieldConfig | Specified spatial and/or spatio-temporal variation in predictors| c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) | c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1)
| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=, Epsilon1=0, Epsilon2=4) | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
| X1_formula | Right-sided formula affecting the 1st linear predictor  | X1_formula = ~ poly(Temp, degree=2 ) | N/A
| X2_formula | Right-sided formula affecting the 2nd linear predictor | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 )

>>>>

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output. While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used. As a result, although $77/80\approx 96\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling. In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method. Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.

@@

In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a large negative impact on the model-based estimates. Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates. In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, with an average decrease by a factir of 0.87.

@@

Of the scenarios without covariates, $10/16\approx63$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates). Adding covariate information resulted in $14/16\approx88$\% of VAST estimates having improved error compared to the stratified mean. The 2 scenarios that produced worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario. The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.

@@

As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is a great deal more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc). 
**For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research**

As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the error in our abundance estimates, while covariate information sometimes lead to an increase in the absolute error between the abundance estimate and true model abundance for Cod and Haddock. More specifically, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, which leads to an incomplete estimation of the covariate response. 
Generally help, but dependent on appropriate covariate response. Yellowtail cases always had appropriate response and thus adding covariates always improved estimates. VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed. (show plot comparing good vs bad response)

increasing temperature scenarios resulted in a larger increase in average error for VAST estimates compared to stratified mean estimates. When including covarites the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constand an dincreasing temperature. Shouldnt model based be more stable with respect to increasing temperature?

@@

<div align="center">Table XX. Settings for our two VAST models.</div>

@@

tabl <- "
|	Parameter	|	Description | Settings A | Settings B
|:--	|:------- |:---- |:----
| *ObsModel* | Link function and assumed distribution | c(10,2) | c(10,2)
|FieldConfig | Specified spatial and/or spatio-temporal variation in predictors| c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) | c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1)
| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=0, Epsilon1=0, Epsilon2=4) | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
| X1_formula | Right-sided formula affecting the 1st linear predictor  |N/A  | X1_formula = ~ poly(Temp, degree=2 )
| X2_formula | Right-sided formula affecting the 2nd linear predictor | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 )

### 2023-04-25 Manuscript/Draft1.tex
a few manuscript updates

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output. While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used. As a result, although \(132/160=82.5\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

Of the scenarios without covariates, \(10/16\approx63\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates). Adding covariate information resulted in \(14/16\approx88\)\% of VAST estimates having improved error compared to the stratified mean. The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario. Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.

Generally help, but dependent on appropriate covariate response. Yellowtail cases always had appropriate response and thus adding covariates always improved estimates. VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed. (show plot comparing good vs bad response)

@@

\textbf{For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without ``perfect'' information? Which covariate is best? Good question for future research}

@@

Table XX. Parameters used in all VAST models.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.14}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.38}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.24}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.24}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Parameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Settings A
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Settings B
\end{minipage} \\
\midrule
\endhead
\emph{ObsModel} & Link function and assumed distribution & c(10,2) & c(10,2) \\
FieldConfig & Specified spatial and/or spatio-temporal variation in predictors & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) \\
RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=, Epsilon1=0, Epsilon2=4) & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
X1\_formula & Right-sided formula affecting the 1st linear predictor & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) & N/A \\
X2\_formula & Right-sided formula affecting the 2nd linear predictor & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) \\
\bottomrule
\end{longtable}

>>>>

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output. While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used. As a result, although \(77/80\approx 96\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

Of the scenarios without covariates, \(10/16\approx63\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates). Adding covariate information resulted in \(14/16\approx88\)\% of VAST estimates having improved error compared to the stratified mean. The 2 scenarios that produced worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario. The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.

@@

As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is a great deal more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).
\textbf{For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without ``perfect'' information? Which covariate is best? Good question for future research}

As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the error in our abundance estimates, while covariate information sometimes lead to an increase in the absolute error between the abundance estimate and true model abundance for Cod and Haddock. More specifically, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, which leads to an incomplete estimation of the covariate response.
Generally help, but dependent on appropriate covariate response. Yellowtail cases always had appropriate response and thus adding covariates always improved estimates. VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed. (show plot comparing good vs bad response)

increasing temperature scenarios resulted in a larger increase in average error for VAST estimates compared to stratified mean estimates. When including covarites the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constand an dincreasing temperature. Shouldnt model based be more stable with respect to increasing temperature?

@@

Table XX. Settings for our two VAST models.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.14}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.38}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.24}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.24}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Parameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Settings A
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Settings B
\end{minipage} \\
\midrule
\endhead
\emph{ObsModel} & Link function and assumed distribution & c(10,2) & c(10,2) \\
FieldConfig & Specified spatial and/or spatio-temporal variation in predictors & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) \\
RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=0, Epsilon1=0, Epsilon2=4) & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
X1\_formula & Right-sided formula affecting the 1st linear predictor & N/A & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) \\
X2\_formula & Right-sided formula affecting the 2nd linear predictor & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) \\
\bottomrule
\end{longtable}
